Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes

### 2023-11-07 Journal-manuscript/Journal-manuscript.tex
Further edits, close to 80% done

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal spatial availability: a singly-constrained
measure of accessibility considering multiple
modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
Antonio Páez\textsuperscript{\getAff{McMaster}}\\
\bigskip
\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
McMaster University, Hamilton, Canada\\
\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
\bigskip
* Corresponding author: soukhoa@mcmaster.ca (AS)\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Recent research has tried to address the way opportunities are counted
in accessibility analysis. In conventional accessibility measures,
opportunities are often multiply counted, which leads to values of
accessibility that are difficult to interpret. Constraining the
calculations to match a known quantity ensures that the measurements sum
up to a predetermined quantity (i.e., the total number of
opportunities), and so each value can be meaningfully related to this
total. A recent such effort is spatial availability, a
singly-constrained accessibility measure. In this paper we extend
spatial availability for use in the case of multiple modes or, more
generally, heterogeneous population segments with distinct travel
behaviors. After deriving a multimodal version of spatial availability
We proceed to illustrate its features using a synthetic example. Next,
we apply it to an empirical example in Madrid. We conclude the paper
with suggestions for future research.

@@

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
from the perspective of planning too {[}see, \emph{inter alia},
4,5--8{]}. Beginning with the work of {[}1{]}, accessibility measures
have been widely used to evaluate the efficiency of transportation
systems when combined with the distribution of opportunities in space
{[}e.g., {]}. As such, it is a holistic measure of spatial systems that
measures the ease of reaching destinations {[}9,10{]}.

In practice, the most common form of accessibility measure is based on
the gravity model. These measures are sums of weighted opportunities
around a focal point (i.e., a potential origin), based on how expensive
it is to reach them. Recent research in accessibility analysis has paid
attention to the way opportunities are counted in the pertinent
calculations. Conventionally, the sums are not constrained, which means
that the same opportunity can enter the sum for different origins.
Counting the same opportunity multiple times treats it as if it was
inexhaustible. But opportunities in general are not inexhaustible, and
in fact some of them are by definition exclusive: for example, once a
job is taken up by someone in the population, the same job is no longer
available for any other person to take. Other types of opportunities are
at the very least subject to congestion: multiple people can avail
themselves of the services of the same family doctor, but the more
people who do, the more congested the service will be.

The issue of congestion was the motivation for the development of
floating catchment area approaches in accessibility analysis
{[}11,12{]}. While these approaches purport to account for congestions,
{[}13{]} demonstrate that in general they do not solve the issue of
multiple counting of opportunities, thus leading to biases in the
calculation of total demand and supply, sometimes inflating them, other
times deflating them. In response to this, recent research has paid
closer attention to the way opportunities are counted in accessibility
analysis. {[}13{]}, for example, tackle floating catchment area methods
and introduce a normalization of the impedance matrix to allocate the
population and then the level of service proportionally. More recently,
{[}14{]} introduced a singly-constrained measure of accessibility,
called spatial availability, that employs a similar, but more
sophisticated proportional allocation mechanism. The work of these
authors shows that floating catchment area methods can be seen as
singly-constrained accessibility measures, and improve on them by
guaranteeing that each opportunity is counted only once - in other
words, treating opportunties as \emph{finite}. The proportional
allocation of spatial availability constrains the calculations to match
a known quantity, therefore ensuring that the measurements sum up to a
predetermined quantity (i.e., the total number of opportunities), and so
each value can be meaningfully related to this total.

A limitation of spatial availability as introduced by {[}14{]} is that
it was developed for the case of a homogeneous population, for example
for the case of a single mode of transportation. However, the finity of
the opportunities makes the analysis of heterogeneous populations very
relevant. In the case of multiple modes of transportation, people who
travel by slow modes (e.g., active modes) can usually reach fewer
opportunities than people who travel by faster modes and whose range is
typically far wider (e.g., car). This implies that slower travelers will
often face increased competition for local opportunities from travelers
who can reach said opportunities from farther afield.

The objective of this paper is to address this limitation of spatial
availability. Our primary motivation is to extend spatial availability
for the case of multimodal accessibility, but this is in fact just one
example of heterogeneous populations (i.e., travel by different modes).
The method itself can easily accommodate other forms of heterogeneity,
for example variations in travel behavior between older and younger
adults {[}e.g., {[}15{]}; {]}, the propensity of older adults to use
different modes of transportation {[}16{]}, the usually shorter trip
lengths of children compared to grown-ups {[}e.g., 17{]}, or the more
limited travel ranges of single parents {[}e.g., 18{]}.

The paper rest of this paper is organized as follows. In Section we use
a synthetic examples with multiple modes to illustrate some relevant
issues; this helps us motivate the derivation of the new spatial
availability expressions. This is followed in Section 3 by an empirical
example using data from the city of Madrid after the implementation of
its Low Emission Zones. Data for this example comes from the city's 2018
travel survey. The example shows the differences in spatial availability
within and outside the LEZ for travellers using different modes, namely
car, transit, cycling and walking. In Section 4, we provide concluding
remarks on the strengths of the use of spatial availability as a
multimodal accessibility measure, and discuss potential future uses in
policy planning scenarios as well as directions for future research.

\hypertarget{a-review-of-multimodal-accessibility-measures}{%
\section{A review of multimodal accessibility
measures}\label{a-review-of-multimodal-accessibility-measures}}

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}19{]}, of which cumulative
opportunity measures and weighted cumulative opportunity measures are
particular forms {[}5{]}. These measures assign a weight to
opportunities based on how easy it is to reach them. Given an origin
(\(i\)) and a destination (\(j\)), an impedance function
\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
generalized cost) into a score that represents the propensity for
interaction. These measures originate from that proposed by {[}1{]},
which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs (\((c_{ij}^m)\)) and/or travel impedance
functions (\(f^m(\cdot)\)).

The Hansen-type measure is not constrained, and as a result does not
consider the opportunities as finite. As an example, the work of
{[}20{]} uses the Hansen-type measure to measure the potential
interaction with retail locations using walking, public transit, and car
modes as \(m\). \(S_i^m\) is the sum of retail locations \(j\) that can
potentially be reached under the travel impedance as calculated for each
\(i\) and \(m\). In other words, each \(i\) has three \(S_i^m\) values,
one per \(m\). In this work, they demonstrate that the car mode has the
highest \(S_i^{m}\) values in the majority of \(i\), i.e., populations
using a car can potentially reach more retail opportunities than
populations using other modes. However, higher \(S_i^{m}\) values for
car do not affect the values of \(S_i^{m}\) for other modes: in effect,
each modes is analyzed as if the others did not exist, which would be
reasonable for an inexhaustible opportunity, but is more questionable
when the opportunity is subject to congestion. Since the measure is
measure is also not constrained, and each opportunity can be and often
is counted multiple times within and between modes, the sum of
accessibility is not a meaningful quantity. To improve interpretability,
\(S_i^m\) values are presented as standardized accessibility index in
this work, from the lowest value of 0 to the highest value of 1. This
presentation or discussion of Hansen-type accessibility measures is
common in the literature {[}e.g., in 21,and 22{]}.

However, this representation of results can make the interpretation of
results challenging \emph{because} values are region-relative and
lacking numeric meaning. Zones will always have values from between 0 to
1 represented, but is a zone with a low value for pedestrian-modes and a
high value for car-modes notable? And if notable, what does the
difference in these standardized values tangibly mean for planners? By
how much should transport systems and land-use configurations be changed
to improve conditions? And in what way can these existing methods be
used to track progress over time? Or between regions? This is
challenging to track, since certain values will always be relatively
`low' and `high'. Alternatively, modal accessibility calculated using
the Hansen-type measure is also presented without standardization, as in
the work by {[}23{]}. Often, the meaning of the accessibility value
itself is not interpreted, as it is difficult to state what having
access to 30,000 employment opportunities by car but having 10,000
employment opportunity by transit tangibly signifies.

However, once opportunities in a region are interpreted as finite, a new
meaning is taken on. As considered in the long tradition of
accessibility work, capacity of opportunities is limited and thus is
subject to competition by population {[}11,14,24--28{]}. There are only
so many school-seats, hospital capacity, employment opportunities, etc.,
in a region and if one person interacts with an opportunity at a given
time, it is taken: the supply of an opportunity and the demand for that
opportunity are two components of accessibility. The competition
consideration is especially important in the case of multiple modes. For
instance, people in a zone who are advantaged with a relatively low
travel-cost mode have the ability to potentially interact with more
opportunities than other people. Due to this advantage, through the
perspective of finite opportunities, there are fewer opportunities left
to be potentially interacted for everyone else, especially zones with
populations using higher travel-cost modes. This recognition is the
motivation behind integrating \emph{competition} for opportunities
within modal accessibility measures.

Arguably, one of the most popular competitive location-based
accessibility measures is the two-step floating catchment area (2SFCA)
approach popularized by {[}12{]} who simplified the approach proposed by
{[}11{]} (with similar considerations for competition in {[}28{]} and
{[}29{]}). The Shen-type accessibility measure's formulation is:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. The Shen-type measure
per mode (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for the \(m\)-using
population in \(i\) over the ravel impedance-weighted demand for
opportunities. In this way, it considers competition but the measure
remains \emph{non-constrained} meaning resulting values are not
associated with global maximums. In other words, it is difficult to
interpret the meaning of differences in Shen-type accessibility scores
between modes in similar ways as reflected for Hansen-type measures.

To illustrate, {[}30{]} calculates \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m={car}\). They demonstrate that
\(i\) with low-income populations have lower \(a_i^m\) than \(i\) with
higher-income populations. Further, they demonstrate that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many \(i\), arguing that this may put \(i\) with lower-income
populations in a further disadvantage. \(a_i\) and/or \(a_i^m\) are used
to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
there is no global maximum, making it is difficult to interpret the
significance between differences in \(a_i^{m}\) values. Questions such
as: what is the impact that competition has on the difference in
\(a_i^m\) values? How does impact vary spatially? And what is the
interpretation of this difference? are left unaddressed.

Spatial availability improves on previously discussed multimodal
accessibility approaches using the Hansen-type measure and the Shen-type
measure by taking into account \emph{competition} in the potential
interaction with opportunities in a \emph{constrained} framework (e.g.,
finite opportunities). This is done by considering: 1) competition
between mass effect (e.g., the advantage of sub-populations residing in
relatively low population-density and high opportunity-proximate areas)
and 2) competition between travel impedance (e.g., sub-populations with
relatively low travel-impedance) through a proportional allocation
mechanism. The following sub-section demonstrates how spatial
availability compares to the Hansen-type and Shen-type measures through
a synthetic example.

\hypertarget{multimodal-spatial-availability-v_im}{%
\subsection{\texorpdfstring{Multimodal spatial availability
\(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}

In brief, we define the spatial availability at \(i\) (\(V_{i}\)) as the
proportion of all opportunities in the region \(O\) that are allocated
to location \(i\) from all opportunity destinations \(j\). \(V_{i}\) is
a value of how many opportunities are available to each location \(i\)
out of all the opportunities in the region. The general formulation of
spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}):

@@

\begin{itemize}
\tightlist
\item
  \(F^t_{ij}\) is a balancing factor that depends on the demand for
  opportunities \(O_j\) and cost of movement in the system
  \(f(c_{ij})\).
\item
  \(V_i\) is the number of spatially available opportunities at \(i\);
  the sum of \(V_{i}\) is equivalent to the total sum of opportunities
  in the region (i.e., \(\sum_j O_j = \sum_i V_i\))
\end{itemize}

The spatial availability measure is introduced in {[}14{]}. Spatial
availability's unique feature is the balancing factor \(F^t_{ij}\), a
proportional allocation mechanisms, that ensures the \(V_i\) calculated
for each \(i\) sums, across all \(i\) in the region, to equal the total
number of opportunities in the region. As such, spatial availability is
a \emph{competitive} and \emph{constrained} accessibility measure as
\(F^t_{ij}\) handles the number of opportunities in the region in a
finite way (proportional allocation). \(F^t_{ij}\) consists of two
components: a population-based balancing factor
\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
respectively, allocates opportunities to \(i\) in proportion to the size
of the population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect).

\(F^p_{i}\) and \$ F\^{}c\_\{ij\}\$ are calculated for each \(i\) such
that they both equal 1 when summed across all \(i\) in the region (e.g.,
\(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
factors are combined multiplicatively to yield \(F^t_{ij}\) which
ensures that a proportion of the opportunities \(O_j\) are allocated to
each \(i\) accordingly. In other words, assuming a finite number of
opportunities in the region, \(F^t_{ij}\) proportionally allocates
\(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
the number of opportunities \emph{spatially available} to the population
at \(i\). Each zonal value is a proportion of the opportunities in the
region (i.e., \(\sum_j O_j = \sum_i V_i\)).

The focus of this paper is to extend \(V_i\) for the measurement of
multimodal accessibility applications. To do so, the balancing factors
are reformulated to yield a proportional value for the set of modes
\(m\) used by populations at each \(i\). As these factors are
proportional, \(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across
each \(m\) at each \(i\) and then across all \(i\) to equal to 1. They
are also similarly combined multiplicatively to obtain their joint
effect, represented as the combined balancing factor \(F^{tm}_{ij}\)
detailed in Equation (\ref{eq:multimodal-balancing-factors}).

@@

\begin{itemize}
\tightlist
\item
  The population balancing factor for each \(m\) at each \(i\) is
  \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
\item
  The cost of travel balancing factor for each \(m\) at \(i\) is
  \(F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}\)
\end{itemize}

Implementing \(F^{tm}_{ij}\), the following Equation
(\ref{eq:spatial-availability-multimodal}) demonstrates the multimodal
configuration of spatial availability \(V_i^m\):

@@

\hypertarget{multimodal-data-and-methods}{%
\subsection{Multimodal data and
methods}\label{multimodal-data-and-methods}}

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal spatial availability: a singly-constrained
measure of accessibility considering multiple
modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
Antonio Páez\textsuperscript{\getAff{McMaster}}\\
\bigskip
\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
McMaster University, Hamilton, Canada\\
\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
\bigskip
* Corresponding author: soukhoa@mcmaster.ca (AS)\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Recent research has aimed to address the way opportunities are counted
in accessibility analysis. In conventional accessibility measures,
opportunities are often multiply counted, which leads to values of
accessibility that are difficult to interpret. Constraining the
calculations to match a known quantity ensures that the measurements sum
up to a predetermined quantity (i.e., the total number of
opportunities), and so each value can be meaningfully related to this
total. A recent such effort is spatial availability, a
singly-constrained accessibility measure. In this paper we extend
spatial availability for use in the case of multiple modes or, more
generally, heterogeneous population segments with distinct travel
behaviors. After deriving a multimodal version of spatial availability
We proceed to illustrate its features using a synthetic example. Next,
we apply it to an empirical example in Madrid. We conclude the paper
with suggestions for future research.

@@

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
from the perspective of planning too {[}see, \emph{inter alia},
4,5--8{]}. Beginning with the work of Hansen {[}1{]}, accessibility
measures have been widely used to evaluate the efficiency of
transportation systems when combined with the distribution of
opportunities in space {[}e.g., {]}. As such, it is a holistic measure
of spatial systems that measures the ease of reaching destinations
{[}9,10{]}.

In practice, the most common form of accessibility measure is based on
the gravity model. These measures are sums of weighted opportunities
around a focal point (i.e., a potential origin), based on how expensive
it is to reach them. Recent research in accessibility analysis has paid
attention to the way opportunities are counted in the pertinent
calculations. Conventionally, the sums are not constrained, which means
that the same opportunity can enter the sum for different origins.
Counting the same opportunity multiple times treats it as if it was
inexhaustible. But opportunities in general are not inexhaustible, and
in fact some of them are by definition exclusive: for example, once a
job is taken up by someone in the population, the same job is no longer
available for any other person to take. More generally, opportunities
are subject to congestion: for example, multiple people can avail
themselves of the services of the same family doctor, but the more
people who do, the more congested the service will be.

The issue of congestion was the motivation for the development of
floating catchment area approaches in accessibility analysis
{[}11,12{]}. While these approaches purport to account for congestions,
Paez et al. {[}13{]} demonstrate that in general they do not solve the
issue of multiple counting of opportunities, thus leading to biases in
the calculation of total demand and supply, sometimes inflating them,
other times deflating them. In response to this, recent research has
paid closer attention to the way opportunities are counted in
accessibility analysis. {[}13{]}, for example, tackle floating catchment
area methods and introduce a normalization of the impedance matrix to
allocate the population and then the level of service proportionally.
More recently, Soukhov et al. {[}14{]} introduced a singly-constrained
measure of accessibility, called spatial availability, that employs a
similar, but more sophisticated proportional allocation mechanism. The
work of these authors shows that floating catchment area methods can be
seen as singly-constrained accessibility measures, and improve on
existing approaches by guaranteeing that each opportunity is counted
only once - in other words, treating opportunities as \emph{finite}. The
proportional allocation of spatial availability constrains the
calculations to match a known quantity, therefore ensuring that the
measurements sum up to a predetermined quantity (i.e., the total number
of opportunities), and so each value can be meaningfully related to this
total.

A limitation of spatial availability as introduced by Soukhov et al.
{[}14{]} is that it was developed for the case of a homogeneous
population, for example for the case of a single mode of transportation.
However, the finite nature of opportunities makes the analysis of
heterogeneous populations very relevant. In the case of multiple modes
of transportation, people who travel by slow modes (e.g., active modes)
can usually reach fewer opportunities than people who travel by faster
modes and whose range is typically far wider (e.g., car). This implies
that slower travelers will often face increased competition for local
opportunities from travelers who can reach said opportunities from
farther afield.

The objective of this paper is to address this limitation of spatial
availability. Our primary motivation is to extend spatial availability
for the case of multimodal accessibility, but it is worthwhile noting
that this is in fact just one case of heterogeneous populations (i.e.,
travel by different modes). The method itself can easily accommodate
other forms of heterogeneity, for example variations in travel behavior
between older and younger adults {[}e.g., 15{]}, the propensity of older
adults to use different modes of transportation {[}16{]}, the usually
shorter trip lengths of children compared to grown-ups {[}e.g., 17{]},
or the more limited travel ranges of single parents {[}e.g., 18{]}.

The paper rest of this paper is organized as follows. In Section we
discuss the use a synthetic examples with multiple modes to illustrate
some relevant issues; this helps us motivate the derivation of the new
spatial availability expressions. This is followed in Section 3 by an
empirical example using data from the city of Madrid after the
implementation of its Low Emission Zones. Data for this example comes
from the city's 2018 travel survey. The example shows the differences in
spatial availability within and outside the LEZ for travelers using
different modes, namely car, transit, cycling and walking. In Section 4,
we provide concluding remarks on the strengths of the use of spatial
availability as a multimodal accessibility measure, and discuss
potential future uses in policy planning scenarios as well as directions
for future research.

\hypertarget{a-brief-review-of-multimodal-accessibility}{%
\section{A brief review of multimodal
accessibility}\label{a-brief-review-of-multimodal-accessibility}}

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}19{]}, of which cumulative
opportunity measures and weighted cumulative opportunity measures are
particular forms {[}5{]}. These measures assign a weight to
opportunities based on how easy it is to reach them. Given an origin
(\(i\)) and a destination (\(j\)), an impedance function
\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
generalized cost) into a score that represents the propensity for
interaction. These measures originate from that proposed by {[}1{]},
which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs (\(c_{ij}^m\)) and/or travel impedance
functions (\(f^m(\cdot)\)).

Hansen-type accessibility is not constrained, which is to say it does
not consider the opportunities as finite. To cite an example, Tahmasbi
et al. {[}20{]} use Hansen-type accessibility to assess the potential
interaction with retail locations by three modes: walking, public
transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
locations \(j\) that can potentially be reached under the travel
impedance as calculated for each \(i\) and \(m\). In other words, for
each origin \(i\) three accessibility scores are calculated. In this
work, Tahmasbi et al. {[}20{]} show that car travel affords the highest
\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
car can potentially reach more retail opportunities than populations
using other modes. However, higher \(S_i^{m}\) values for car do not
affect the values of \(S_i^{m}\) for other modes: in effect, each mode
is analyzed as if the others did not exist. Since the measure is not
constrained, each opportunity is typically counted multiple times within
and between modes, and as a result the sum of accessibility is not
necessarily a meaningful quantity. The accessibility scores for the
modes are often values that are difficult to interpret beyond making
statements about relative size. For example, Lunke {[}21{]}, researching
the region of Oslo, reports accessibility scores for car in the order of
tens of thousands of employment opportunities. The corresponding scores
for transit are lower, but still often in the thousands or tens of
thousands. As reported, the ratio of the transit to the car score can be
lower than 0.2 (meaning transit gives access to less than 20\% of the
opportunities than car). But despite the discussion about ``sufficient
accessibility'' (e.g., in the abstract), it is unclear what the
unconstrained scores mean: is having access to but 10,000 jobs by
transit insufficient? After all, 10,000 employment opportunities are
still plenty of opportunities. These ratios can be found elsewhere in
the literature {[}e.g., 22, and 9,15,18, and 8{]}, and they are useful
to assess when some members of the public are better or worse off than
others, but they do not say much about how bad is ``worse''.

Besides ratios of accessibility, another way to improve interpretability
of scores sometimes seen in the literature is to standardize them to lie
in the range {[}0-1{]}. This adjustment is only helpful insofar as it
facilitates relative comparisons, but interpretation of the scores
remains challenging because the values are specific to a region and
convey no meaning about the magnitude of the scores. In this approach,
zones always have values between 0 and 1, but how remarkable is a zone
with a low score for pedestrians and a high value for car? And if
remarkable, what does the difference in these standardized values mean
for planners? By how much should transport systems and land-use
configurations be changed to improve conditions? And in what way can
these scores be used to track differences over time? Or between regions?
These questions lack straightforward answers since certain values will
always be relatively `low' or `high', but do not track to a quantity
that can be intuitively understood. Presentation or discussion of
Hansen-type accessibility that has been standardized in this way is not
uncommon in the literature {[}e.g., 23,24{]}.

Once we understand opportunities to be finite, it is possible for an
accessibility measure to take a crisper meaning. As considered in the
long tradition of accessibility research, capacity of opportunities is
limited and thus is subject to competition by population
{[}11,14,25--29{]}. There are only so many school-seats, hospital
capacity, employment opportunities, etc., in a region and if one person
reaches an opportunity, it is taken: the supply of an opportunity and
the demand for that opportunity are two components of accessibility.
These are clear examples of opportunities that are clearly competitive.
But we would go as far as to argue that every type of opportunity is
subject to congestion or capacity constraint, even when the
opportunities are conventionally seen as non-competitive. Amenities are
a good example of this. For instance, standards for providing green
spaces are often stated in the form of \emph{exclusive access}, in units
of amenity per capita. A case in point is Natural England, an
organization that recommends an Accessible Natural Greenspace Standard
such that the minimum supply of space is one ha of statutory Local
Nature Reserves per thousand population\footnote{see
  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
Similarly, the World Health Organization {[}cited in 30{]} recommends
that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
For our purposes, standards of this type translate into ``how much of a
resource can be offered to each individual that is not allocated to
anyone else?''. Green spaces often have large capacities, but they still
have a capacity. Constraining accessibility is in this way a useful way
to evaluate the congested availability of any type of opportunity. As
development of sound standards is emphasized in the planning literature,
in particular in regards to fairness in transportation {[}see 31{]},
spatial availability analysis is a useful way to develop and assess
standards.

The relevance of the considerations above is put in sharper relief when
we think about the use of multiple modes (or heterogeneous populations).
If we return to Oslo for a moment {[}21{]}, we notice that the places
that have high accessibility by transit are also the places that have
\emph{very high} accessibility by car (Figure 2). Those two populations
are going for the same opportunities, and those travelling by transit
have fewer to choose from to begin with. More generally, people in a
zone who are advantaged with relatively cost of travel will have the
ability to potentially reach more opportunities than other people. Due
to this advantage, through the perspective of finite opportunities,
there are fewer opportunities left for everyone else, especially for
those who use modes that are slower or more expensive.

As noted in the introduction, competitive accessibility was the
rationale for developing floating catchment area methods (FCA),
popularized by {[}12,11,although similar, and earlier, developments are
found in 29,32{]}. Shen-type accessibility is formulated as:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. Shen-type modal
accessibility (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
the travel impedance-weighted demand for opportunities. In this way, it
considers competition. That said, the measure remains unconstrained,
meaning both population \emph{and} opportunities are multiply counted
{[}see 13{]}. In other words, interpretation of differences in Shen-type
accessibility scores between modes is fraught as it is for Hansen-type
measures.

To illustrate, {[}33{]} calculate \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m=\text{car}\). Their results
indicate that zones with low-income populations have lower \(a_i^m\)
than zones with higher-income populations. Further, they show that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many zones, arguing that this may further place those zones with
lower-income populations at disadvantage. \(a_i\) and/or \(a_i^m\) are
used to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
opportunities were doubly counted (entering the sums of both modes),
this makes for uneasy interpretations of the differences in \(a_i^{m}\)
between modes. Questions that this approach leaves unaddressed include:
what is the impact of competition on the difference in \(a_i^m\) values?
How does the impact vary spatially? And what is the interpretation of
this difference?

Spatial availability improves on previously discussed accessibility
approaches using the Hansen-type measure and the Shen-type measure by
constraining the sum of opportunities, that is, by treating
opportunities as finite. This is done by means of proportional
allocation factors that follow well established principles of spatial
interaction and the gravity model {[}see 34{]}. In Soukhov et al.
{[}14{]} these factors consider: 1) the mass effect (e.g., the size of
populations at different origins); and 2) the cost of travel from
different zones (e.g., some sub-populations face relatively higher or
lower costs). The following section introduces the multimodal form of
spatial availability.

\hypertarget{multimodal-spatial-availability-v_im}{%
\section{\texorpdfstring{Multimodal spatial availability
\(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}

In brief, we define the spatial availability at \(i\) (\(V_{i}\)) as the
proportion of all opportunities in the region \(O\) that are allocated
to location \(i\) from all opportunity destinations \(j\). \(V_{i}\) is
a value of how many opportunities are available to each location \(i\)
out of all the opportunities in the region. The general formulation of
spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}) {[}see 14{]}:

@@

\begin{itemize}
\item
  \(F^t_{ij}\) is a balancing factor that depends on the size of the
  populations at different locations that demand opportunities \(O_j\),
  as well as the cost of movement in the system \(f(c_{ij})\).
\item
  \(V_i\) is the number of spatially available opportunities at \(i\);
  the sum of \(V_{i}\) is identical to the total number of opportunities
  in the region (i.e., \(\sum_j O_j = \sum_i V_i\)); in other words,
  opportunities are dealt with as finite resources.
\end{itemize}

Compared to Hansen-type accessibility:

\begin{equation}
\label{eq:hansen-accessibility}
A_i = \sum_{j=1}^J O_jf(c_{ij})
\end{equation}

\noindent we see that spatial availability is, like Hansen's measure, a
weighted sum of the opportunities. What makes spatial availability stand
apart from other approaches is the weight used in the sum, balancing
factor \(F^t_{ij}\), implements a proportional allocation mechanisms to
ensure that the sum of \(V_i\) is constrained to match the total number
of opportunities in the region. As such, spatial availability is
singly-constrained and naturally implements competition or congestion.
\(F^t_{ij}\) consists of two parts. The first part is a population-based
proportional allocation factor to model the mass effect of the gravity
model: \[
F^p_{i} = \frac{P_i}{\sum_i P_i}
\]

This factor makes opportunities available based on demand. Secondly,
there is an impedance-based proportional allocation factor that model
the cost effect: \[
F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}
\]

This factor makes opportunities available preferentially to those who
can reach them at a lower cost. \(F^p_{i}\) and \(F^c_{ij}\) are
designed so that they both equal 1 when summed across all \(i\) in the
region (e.g., \(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These
factors are combined multiplicatively to yield \(F^t_{ij}\) which
ensures that a proportion of the opportunities \(O_j\) are allocated to
each \(i\) accordingly. In other words, assuming a finite number of
opportunities in the region, \(F^t_{ij}\) proportionally allocates
\(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
the number of opportunities \emph{available} to the population at \(i\).
Each zonal value is a proportion of the opportunities in the region
(i.e., \(\sum_j O_j = \sum_i V_i\)).

The focus of this paper is to extend \(V_i\) for the measurement of
multimodal applications (or more generally heterogeneous populations).
To do so, the balancing factors need to be reformulated so that 1) the
mass effect now accounts not only for the size of the population at
\(i\), but also the size of sub-populations within \(i\); and 2) the
cost of travel is not only for different zones, but by sub-populations
within each zone (e.g., the cost of travel from \(i\) by car, transit,
walking, etc.) When we introduce modes (or sub-populations) \(m\), the
proportional allocation factors need to satisfy the condition that
\(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across each \(m\) at
each \(i\) and then across all \(i\) to equal to 1. They are also
similarly combined multiplicatively to obtain their joint effect,
represented as the combined balancing factor \(F^{tm}_{ij}\) similar to
that detailed in Equation (\ref{eq:multimodal-balancing-factors}). This
factor is given by:

@@

\noindent where:

\begin{itemize}
\tightlist
\item
  The factor for allocation by population for each \(m\) at each \(i\)
  is \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\); and
\item
  The factor for allocation by cost of travel for each \(m\) at \(i\) is
  \(F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}\)
\end{itemize}

Implementing \(F^{tm}_{ij}\), the following Equation
(\ref{eq:spatial-availability-multimodal}) gives the multimodal version
of spatial availability \(V_i^m\):

@@

\hypertarget{data-and-methods}{%
\subsection{Data and methods}\label{data-and-methods}}
